---
title: 'spreadr: A R package to simulate spreading activation in a network'
author: "Cynthia S. Q. Siew"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

<style>
p { text-align: justify }
</style>

# Introduction

The notion of spreading activation is a prevalent metaphor in the cognitive sciences; however, the tools to implement spreading activation in a computational simulation are not as readily available.
This vignette introduces the spreadr R package (pronunced 'SPREAD-er'), which can implement spreading activation within a specified network structure.
The algorithmic method implemented in `spreadr` subroutines followed the approach described in Vitevitch, Ercal, and Adagarla (2011), who viewed activation as a fixed cognitive resource that could “spread” among connected nodes in a network.
See Vitevitch et al. (2011) for more details on the implementation of the spreading activation process.

# Installation {.tabset}

You can choose to install the stable version from CRAN, or the development build from GitHub.

## Stable (CRAN)

```{r eval=FALSE}
install.packages("spreadr")
```

If you encounter any bugs or issues, please try the development build first.
The bug or issue may have already been fixed on GitHub, but not yet propagated onto CRAN.

## Development (GitHub)

```{r eval=FALSE}
install.packages("remotes")
remotes::install_github("csqsiew/spreadr")
```

If you encounter any bugs or issues, please try the development build first.
The bug or issue may have already been fixed on GitHub, but not yet propagated onto CRAN.

# Example with a phonological network

In this example, we will simulate spreading activation in a small sample portion of a phonological network (Ying and Vitevitch 2009) which is automatically loaded with spreadr.
This phonological network is unweighted and undirected, but spreadr supports weighted and directed graphs as well.
This makes it is possible to simulate spreading activation in a weighted network where more activation is passed between nodes that have "stronger" edges, or in a directed (asymmetric) network where activation can pass from node *i* to node *j* but not necessarily from node *j* to node *i*.

You may substitute any network in place of this example one.

## Describe the network {.tabset}

The network for spreading activation must be either an `igraph` object or an adjacency matrix.

### Using an igraph object

`pnet`, an `igraph` object with named vertices representing our sample phonological network is automatically loaded with the spreadr library.

```{r message=FALSE}
library(spreadr)
library(igraph)

pnet  # inspect the igraph object
```

```{r eval=FALSE}
plot(pnet)
```
<!-- Behind the scenes, we want to generate identical plots, so save a layout.
     This is hidden because I don't want to make the vignette too complex. -->
```{r echo=FALSE}
set.seed(1)
layout <- layout_with_fr(pnet)
plot(pnet, layout=layout)
```

Don't worry if your plot does not look exactly as above.
The layout of the nodes is determined stochastically.

### Using an adjacency matrix

`pnetm`, an named adjacency matrix representing our sample phonological network is automatically loaded with the spreadr library.

```{r message=FALSE}
library(spreadr)
library(igraph)
set.seed(1)

pnetm[1:5, 1:5]  # inspect the first few entries
```

```{r eval=FALSE}
plot(graph_from_adjacency_matrix(
  pnetm, mode="undirected")) # visualise the graph
```
```{r echo=FALSE}
plot(graph_from_adjacency_matrix(
    pnetm, mode="undirected"),
  layout=layout)
```

For those following along: don't worry if your plot does not look exactly as above.
The layout of the nodes is determined stochastically.

For simplicity, the rest of this example will use `pnet`, an `igraph` representation of `pnetm`.
This difference is trivial --- the spreadr function accepts both igraph and adjacency matrix.

## Specify the adding of activation {.tabset}

A simulation of spreading activation is uninteresting without some activation to spread.
In the simplest case, you may specify the initial activation state of each node, from which the simulation will proceed.
Alternatively, you can also specify the addition of activation at any node, at any time point within the simulation.

### Initial activation only

Initial activation is specified by a `data.frame` (or `data.frame`-like object, such as a `tibble`) with columns `node` and `activation`.
Each row represents the addition of `activation` amount of activation to the node with name `node`, at the initial pre-spreading activation step of the simulation.

For example, if we wanted the nodes `"beach"` and `"speck"` to have `20` and `10` activation initially, respectively:

```{r}
start_run <- data.frame(
  node =       c("beach", "speck"),
  activation = c(     20,      10))
```

### Activation at specified time points

The adding of activation at specified time points is specified by a `data.frame` (or `data.frame`-like object, such as a `tibble`) with columns `node`, `activation`, and `time`.
Each row represents the addition of `activation` amount of activation to the node with name `node` at time point `time`.
The initial state before any spreading activation occurs is `time = 0`.

Therefore, if we wanted the node `"beach"` to have `20` activation initially, then `"speck"` to have `10` activation at `time = 5`:

```{r}
start_run <- data.frame(
  node =       c("beach", "speck"),
  activation = c(     20,      10),
  time =       c(      0,       5))
```

For more details about the meaning of `time` in the spreading activation simulation, please read the `spreadr` function documentation.
To keep things simple, the rest of this example will use the version of `start_run` without the `time` column (see the leftmost tab).

```{r echo=FALSE}
start_run <- data.frame(
  node =       c("beach", "speck"),
  activation = c(     20,      10))
```

## Run the simulation

We are now ready to run the simulation through the `spreadr` function.
This function takes a number of parameters, some of which are listed here.
Required parameters are written in **bold**, optional parameters are written with their default values (in brackets):

i. **network**: Adjacency matrix or `igraph` object representing the network in which to simulate spreading activation
ii. **start_run**: Non-empty `data.frame` describing the addition of activation into the network.
iii. *decay* (`0`): Proportion of activation lost at each time step (range from 0 to 1)
iv. *retention* (`0.5`): Proportion of activation retained in the originator node (range from 0 to 1)
v. *suppress* (`0`): Nodes with activation values lower than this value will have their activations forced to `0` (typically this will be a very small value e.g. < 0.001)
vi. *time* (`10`): Number of iterations in the simulation
vii. *include_t0* (`FALSE`): If the initial state activations should be included in the result `data.frame`

For more details on all the possible function parameters, please read the `spreadr` function documentation.

```{r}
result <- spreadr(network=pnet, start_run=start_run, include_t0=TRUE)
```

## Examine the results

The result is presented as a `data.frame` with columns `node`, `activation`, and `time`.
Each row contains the activation value of a node at its specified time step.

```{r eval=TRUE}
head(result)  # inspect the first few rows
tail(result)  # inspect the last few rows
```

As a `data.frame`, the `result` can be easily saved as a CSV file for sharing.

```{r eval=FALSE}
write.csv(result, file="result.csv")  # save result as CSV file
```

You can also easily visualise the `result` using your favourite graphical libraries.
Here, we use `ggplot2`:

```{r}
library(ggplot2)
ggplot(result, aes(x=time, y=activation, color=node)) +
  geom_point() +
  geom_line()
```

# Additional examples

This section shows how spreadr can be implemented in a weighted network (i.e., the edges in the network can have different weights) and a directed network (i.e., the edges in the network are not symmetric).

## Using a weighted network

## Using a directed network

## Repeating the simulation across many parameters

## Creating an animation of spreading activation

(i) Weighted, undirected network

```{r eval=FALSE}
# weighted, undirected network example
g_w <- g
set.seed(2)
E(g_w)$weight <- runif(gsize(g_w)) # make the edges in the network have different weights ranging from 0 to 1 (excluding extreme values)

result_w <- spreadr::spreadr(start_run = initial_df, decay = 0,
                              retention = 0.5, suppress = 0,
                              network = g_w, time = 10)

result_w_t0 <- rbind(result_w, a1) # add back initial activation at t = 0
ggplot(data = result_w_t0, aes(x = time, y = activation, color = node, group = node)) +
  geom_point() + geom_line() + ggtitle('weighted, undirected network') # visualize the results
```

As you can see from the plot, the result is slightly different from the original example with unweighted edges. More activation is passed to the node that is more "strongly" connected to the originator node.

(ii) Unweighted, directed network

For this example, we will create the network via an adjacency matrix.

```{r eval=FALSE}
# unweighted, directed network example
set.seed(3)
g_d_mat <- matrix(sample(c(0,1), 100, replace = T), 10, 10) # make a matrix and randomly fill some cells with 1s
diag(g_d_mat) <- 0 # remove self-loops
g_d_mat

# spreadr will work on an adjacency matrix too
result_d <- spreadr::spreadr(start_run = data.frame(node = 1, activation = 20, stringsAsFactors = F),
                             decay = 0,
                              retention = 0.5, suppress = 0,
                              network = g_d_mat, time = 10)

result_d_t0 <- rbind(result_d, data.frame(node = 1, activation = 20, time = 0)) # add back initial activation at t = 0
ggplot(data = result_d_t0, aes(x = time, y = activation, color = node, group = node)) +
  geom_point() + geom_line() + ggtitle('unweighted, directed network')
```

The (_i_, _j_) entry of the adjacency matrix should represent the edge going from source node _i_ to target node _j_.
See below and compare it to the plot above.

```{r eval=FALSE}
data.frame(node = 1:10, edge = g_d_mat[,1])
data.frame(node = 1:10, edge = g_d_mat[1,])
```

# Reporting bugs or issues

First, please check that you are on the development build.
This is because the development build contains a more up-to-date version of spreadr, which may include various bug fixes, as compared to the stable CRAN build.

If your problems persist, please report them on our [Github repository](https://github.com/csqsiew/spreadr/issues).

# References

Siew, C. S. Q. (under review). spreadr: A R package to simulate spreading activation in a network.
Vitevitch, M. S., Ercal, G., & Adagarla, B. (2011). Simulating retrieval from a highly clustered network: Implications for spoken word recognition. *Frontiers in Psychology, 2*, 369.
